Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-04 submission.tex
Added some results from tag set experiments

\cite{Meg:01,Meg:02} trained and evaluated a range of PoS taggers on the
Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a tag set based
on a Swedish version of PAROLE tags totaling 139 tags. Furthermore, they
investigated the effects of tag set size on tagging by mapping the original tag
set into smaller subsets designed for parsing. They argue that a tag set with
complete morphological tags may not be necessary for all NLP applications, for
instance syntactic parsing. They found that the smallest tag set comprising 26
tags yields the lowest tagger error rate. However, for some of the taggers,
augmenting the tag set with more linguistically informative tags may actually
lead to a drop in error rate. They argue that this shows that the size of the
tag set as well as the type of information in the tags are crucial factors for
tagger performance. However/unfortunately, they do not report results of
parsing with the various PoS tag sets.

@@

\paragraph{Baseline}
It is common practice to compare the performance of PoS taggers to a
pre-computed baseline for an initial point of comparison.
For PoS tagging, a commonly used baseline is the Most Frequent Tag (MFT)
baseline, which we use in our experiments. This involves labeling each word
with the tag it was assigned most frequently in the training. All unknown
words, i.e., words not seen in the training data, are assigned the tag most
frequently assigned to words seen only once in the training. Unknown and
infrequent words have in common that they rarely occur, and we might therefore
expect them to have similar properties.
%, there are two main
%approaches: assign it the most frequent tag overall in the training data, or
%the tag most frequently assigned to words seen only once in the training data.
%We adopted the latter approach for our baseline tagger, as we believe that
%unknown words may have much in common with words that occur only once in the
%training, more so than simply words of the most frequent part-of-speech. The
%reason for this is that unknown and infrequent words have in common that they
%rarely occur, and we might therefore expect them to have similar properties.

@@

The morphological properties of the various parts-of-speech is reflected in the
morphological features associated with the respective PoS tags. In addition to
morphological properties such as definiteness, tense and number, all classes
except for verbs can be distinguished "on" type, e.g., common nouns and proper
nouns, while pronouns can be reflexive, reciprocal, personal or interrogative.

\subsection{Experiments}
\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine, neuter),
definiteness (indefinite or definite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners). Additionally, NDT has a separate case
feature to distinguishing nouns in genitive case.

@@

@@

@@

@@

@@

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Tag Set} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & \textbf{87.15\%} & \textbf{90.39\%} \\
        %\bottomrule
    %\end{tabular}
    %\caption{Evaluation of tagging and parsing the development data with the two
        %initial tag sets.}
    %\label{inittagseteval}
%\end{table}

>>>>

\newcite{Meg:01} and \newcite{Meg:02} trained and evaluated a range of PoS
taggers on the Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a
tag set based on a Swedish version of PAROLE tags totaling 139 tags.
Furthermore, they investigated the effects of tag set size on tagging by
mapping the original tag set into smaller subsets designed for parsing. They
argue that a tag set with complete morphological tags may not be necessary for
all NLP applications, for instance syntactic parsing. They found that the
smallest tag set comprising 26 tags yields the lowest tagger error rate.
However, for some of the taggers, augmenting the tag set with more
linguistically informative tags may actually lead to a drop in error rate. They
argue that this shows that the size of the tag set as well as the type of
information in the tags are crucial factors for tagger performance.
However/unfortunately, they do not report results of parsing with the various
PoS tag sets.

@@

\paragraph{Baseline}
It is common practice to compare the performance of PoS taggers to a
pre-computed baseline for an initial point of comparison.
For PoS tagging, a commonly used baseline is the Most Frequent Tag (MFT)
baseline, which we use in our experiments. This involves labeling each word
with the tag it was assigned most frequently in the training. All unknown
words, i.e., words not seen in the training data, are assigned the tag most
frequently assigned to words seen only once in the training. Unknown and
infrequent words have in common that they rarely occur, and we might therefore
expect them to have similar properties.

@@

The morphological properties of the various parts-of-speech is reflected in the
morphological features associated with the respective PoS tags. For instance,
as nouns can take on gender, definiteness and number, additionally "being of"
genitive case, the treebank operates with features for gender, definiteness,
number and case with "accompanying" values. In addition to
morphological properties such as definiteness, tense and number, all classes
except for verbs can be distinguished "on" type, e.g., common nouns and proper
nouns, while pronouns can be reflexive, reciprocal, personal or interrogative.

\subsection{Experiments}
"Nearly" all tag set modifications leads to a drop in tagger accuracy, as
expected. Few exceptions...

\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine, neuter),
definiteness (indefinite or definite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners). Additionally, NDT has a separate case
feature to distinguishing nouns in genitive case.

@@

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        %Case & 93.77\% & \textbf{97.48\%} & 87.63\% & 90.72\% \\
        %Definiteness & 89.67\% & 97.00\% & 88.27\% & 91.42\% \\
        %Gender & 89.54\% & 96.09\% & 87.21\% & 90.36\% \\
        %Number & 90.04\% & 96.37\% & 87.97\% & 91.00\% \\
        %Type & 91.90\% & 96.92\% & 88.07\% & 91.11\% \\
        %Case \& definiteness & 89.65\% & 97.03\% & 88.39\% & 91.44\% \\
        %Type \& case & 91.73\% & 96.92\% & 88.46\% & 91.51\% \\
        %Type \& definiteness & 89.65\% & 96.99\% & 88.44\% & 91.48\% \\
        %Type \& number & 90.02\% & 96.37\% & 87.95\% & 90.95\% \\
        %Type, case \& definiteness & 89.61\% & 97.05\% & \textbf{88.81\%} &
        %\textbf{91.73\%} \\
        %Type, definiteness \& number & 88.39\% & 96.46\% & 88.06\% & 91.12\%
        %\\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for nouns.}
    %\label{substresults}
%\end{table}

@@

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Mood & 94.12\% & 97.43\% & 87.04\% & 90.19\% \\
        %Tense & 93.74\% & 97.30\% & 86.97\% & 90.18\% \\
        %Voice & 94.13\% & 97.45\% & 86.96\% & 90.09\% \\
        %Mood \& tense & 93.74\% & 97.31\% & 87.12\% & 90.31\% \\
        %Voice \& tense & 93.74\% & 97.28\% & 86.99\% & 90.15\% \\
        %Mood, tense \& voice & 93.74\% & 97.27\% & 86.83\% & 90.05\% \\
        %Finiteness & 93.72\% & 97.35\% & \textbf{87.30\%} &
        %\textbf{90.43\%} \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for verbs.}
    %\label{verbresults}
%\end{table}

@@

All features except for number (LAS of 86.99\%) led to increases in parser
accuracy scores. Degree being the most promising with a reported LAS of
87.29\%, while distinguishing adjectives on definiteness yield an LAS of
87.14\% and gender leads to LAS of 87.10\%.

Combining the most promising features, we found that none of the combinations
surpass the LAS with the distinction of degree alone. Definiteness and degree
with an LAS of 87.23\% and definiteness and number with an LAS of 87.27\% and
UAS identical to that of degree alone. The most fine-grained distinction,
definiteness, degree and number, yields an LAS of 87.14\%.

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Definiteness & 93.45\% & 96.84\% & 87.14\% & 90.29\% \\
        %Degree & 94.13\% & 97.41\% & \textbf{87.29\%} & \textbf{90.44\%} \\
        %Gender & 93.56\% & 96.89\% & 87.10\% & 90.25\% \\
        %Number & 93.51\% & 96.71\% & 86.99\% & 90.10\% \\
        %Type & 94.12\% & 97.40\% & 87.11\% & 90.25\% \\
        %Definiteness \& degree & 93.45\% & 96.81\% & 87.23\% & 90.39\% \\
        %Definiteness \& gender & 92.94\% & 96.31\% & 87.18\% & 90.39\% \\
        %Definiteness \& number & 93.48\% & 96.78\% & 87.27\% &
        %\textbf{90.44\%} \\
        %Degree \& gender & 93.56\% & 96.87\% & 87.00\% & 90.16\% \\
        %Degree \& number & 93.49\% & 96.76\% & 87.13\% & 90.26\% \\
        %Definiteness, degree \& number & 93.47\% & 96.81\% & 87.14\% & 90.30\% \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for adjectives.}
    %\label{adjresults}
%\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Feature} & \textbf{MFT} & \textbf{Accuracy} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        Definiteness & 94.13\% & 97.49\% & \textbf{87.30\%} &
        \textbf{90.42\%} \\
        Gender & 93.86\% & 97.28\% & 87.09\% & 90.31\% \\
        Number & 94.06\% & 97.49\% & 87.04\% & 90.18\% \\
        Type & 94.14\% & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for determiners.}
    \label{detresults}
\end{table}

@@

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        %Case & 94.12\% & 97.50\% & 87.08\% & 90.21\% \\
        %Gender & 94.13\% & 97.48\% & 87.06\% & 90.23\% \\
        %Number & 94.13\% & 97.49\% & 87.21\% & 90.33\% \\
        %Person & 94.14\% & 97.49\% & 87.22\% & 90.32\% \\
        %Type & 94.14\% & 97.48\% & 87.19\% & 90.40\% \\
        %Number \& person & 94.13\% & 97.49\% & 96.98\% & 90.16\% \\
        %Type \& case & 94.12\% & 97.51\% & \textbf{87.30\%} &
        %\textbf{90.41\%} \\
        %Type \& number & 94.13\% & 97.49\% & 87.27\% & \textbf{90.41\%} \\
        %Type \& person & 94.14\% & 97.49\% & 87.00\% & 90.14\% \\
        %Type, case \& number & 94.12\% & \textbf{97.52\%} & 87.11\% & 90.36\% \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for pronouns.}
    %\label{pronresults}
%\end{table}
